---
title: "Survey Weighting Exercise: Nonvoter Study Analysis"
date: "2025-08-21"
author: "Derek Willis/Rob Wells"
editor_options: 
  markdown: 
    wrap: 72
---

**YOUR NAME HERE: Princess Tatsi**

Activity: Belonging Survey Methods

Let's revisit the summary of UMD's Belonging Survey, and look at the
appendix in particular. <https://president.umd.edu/internal/belonging>

Working in pairs or small groups, each person reads the appendix closely
and identifies any statements that raise methodological concerns or are
unclear. Discuss your findings, summarize your discussion and then
submit no more than three most important concerns or questions you have
and how they might impact the results of the Belonging Survey.

**Write your answer here:** 1. The first concern is that racial/Ethnic
groups with low response rates were collaspsed into one group because
"none of them had enough

2.  Responses for those who identified as "genderfluid, non-binary,
    transgender, and/or wrote in another gender identity" were collapsed
    into the category gender expansive

3.  "Over 25,000 qualitative responses were analyzed using content
    analysis to extract common themes and thematic patterning across
    respondents."

#-------------------------------------------------------------------------------
#Survey Weighting
#-------------------------------------------------------------------------------

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)

```

# Introduction

This notebook explores how survey weighting affects data analysis and
interpretation in journalism. We'll use a dataset from a 2020 survey on
voting behavior to understand weighting concepts and how they can change
our understanding of public opinion data. The codebook with the
questions from the survey is here:
<https://github.com/dwillis/jour405_files/blob/main/nonvoters_codebook.pdf>

## What is Survey Weighting?

Survey weighting is like adjusting the volume on different voices in a
conversation to make sure everyone is heard equally. Think of it like
running a mixing board at a concert. In survey research:

-   Some groups of people are more likely to answer surveys than others
-   This can make the survey sample different from the actual population
-   Survey weights help fix this imbalance by giving more "volume" to
    underrepresented groups and less to overrepresented groups

**For example:** If your survey has too many college graduates compared
to the general population, you might give each college graduate's
response a weight of 0.8 (counting as 80% of a response) and each
non-college graduate a weight of 1.2 (counting as 120% of a response).

## Why Should Journalists Care?

-   Unweighted results might give a misleading picture of public opinion
-   Proper weighting helps ensure your reporting accurately represents
    the community
-   Understanding weighting helps you critically evaluate polls from
    other sources
-   Small differences in weighting can sometimes change the headline of
    your story

## Our Questions

1.  How does weighting affect our estimates of voting intentions?
2.  Which demographic groups show the largest differences between
    weighted and unweighted results?
3.  How might weighting influence reporting of this data?

# Data Overview

The nonvoters dataset contains survey responses about voting intentions
and behaviors from 5,836 American adults. The survey was conducted prior
to the 2020 election and includes a variety of demographic information
and political attitudes.

```{r load-data}
# Load the dataset
nonvoters_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/nonvoters_data.csv")

# Look at the structure of the data
glimpse(nonvoters_data)
```

Let's break down the key variables we'll focus on:

-   `weight`: The survey weight assigned to each respondent, showing how
    much their response "counts"
-   `Q21`: Voting intention with these values:
    -   1 = Yes (plans to vote)
    -   2 = No (does not plan to vote)
    -   3 = Unsure/Undecided
-   `voter_category`: Respondent's voting history:
    -   "always": Votes in all or nearly all elections
    -   "sporadic": Votes occasionally
    -   "rarely/never": Rarely or never votes
-   Demographics: `race`, `gender`, `educ` (education level),
    `income_cat`

# Understanding Survey Weights

Before we dive into the analysis, let's examine the weights in our
dataset to understand how they're distributed.

```{r examine-weights}
# Summary statistics of weights
summary(nonvoters_data$weight)
```

```{r}
# Histogram of weights
ggplot(nonvoters_data, aes(x = weight)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of Survey Weights",
    x = "Weight",
    y = "Count"
  )
```

**Explanation:** - The summary statistics show us the minimum, maximum,
and typical weights in the dataset - The histogram shows how many
respondents have each weight value - Most weights cluster around 1.0
(which means those respondents count as exactly one person) - Weights
below 1.0 mean a respondent is overrepresented (counts as less than one
person) - Weights above 1.0 mean a respondent is underrepresented
(counts as more than one person)

**Questions:**

1.  What is the range of weights in this dataset? **Write your answer
    here:** The range of weights is 2.8

The mean and median are almost close which indicates an almost
symmetrical distribution with no extreme outliers.

2.  What does it mean when a respondent has a weight of 0.23? What about
    3.04? **Write your answer here:** A respondent with a weight of 0.23
    is underrepresented by the data so counts as more than one person.
    And a person weighed at 3.04 is overrepresented by the data so they
    count as less than a person

3.  Why might some respondents be weighted more heavily than others?
    **Write your answer here:** Some respondents will be weighed more
    than others to account for their underrepresentation in the actual
    dataset.For example, polling results did not accurately project the
    number of registered Trump voters

# Exercise 1: Voter Categories and Weighting

First, let's look at how many people fall into each voting category in
our sample.

```{r voter-categories}
library(knitr)
# Count of respondents by voter category 
nonvoters_data |>
  count(voter_category) |>
  mutate(percentage = n / sum(n) * 100) 
```

**Explanation:** This table shows the breakdown of our sample by voting
history. We can see that most respondents (44.1%) are "sporadic" voters,
followed by "always" voters (31.0%), and "rarely/never" voters (24.9%).

Now, let's compare unweighted and weighted voting intentions by voter
category.

## Part A: Unweighted Analysis

First, we'll calculate the raw percentages without applying weights:

```{r unweighted-by-category}
# Calculate unweighted percentages
unweighted_by_voter_category <- nonvoters_data |>
  # Filter out missing responses
  filter(!is.na(Q21), Q21 > 0) |>
  # Group by voter category and response
  group_by(voter_category, Q21) |>
  # Count responses in each group
  summarize(count = n(), .groups = "drop_last") |>
  # Calculate percentages within each voter category
  mutate(total = sum(count),
         percentage = count / total * 100) |>
  ungroup()

# Create a more readable version with voting intentions as columns
unweighted_summary <- unweighted_by_voter_category |>
  pivot_wider(
    id_cols = voter_category,
    names_from = Q21,
    values_from = percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

unweighted_summary
```

**Explanation:** - This table shows the percentage of respondents in
each voter category who said "Yes," "No," or "Unsure" when asked if they
plan to vote - We calculated these percentages by simply counting
responses and dividing by the total in each category - No weights have
been applied yet—each respondent counts equally - Notice that even among
"rarely/never" voters, a high percentage say they plan to vote in the
upcoming election

## Part B: Weighted Analysis

Now, let's apply the survey weights to the same analysis. Instead of
just counting respondents, we'll sum their weights:

```{r weighted-by-category}
# YOUR TASK: Calculate the weighted percentages
# Use the weight variable to adjust the counts

weighted_by_voter_category <- nonvoters_data |>
  # Filter out missing responses, same as before
  filter(!is.na(Q21), Q21 > 0) |>
  # Group by voter category and response
  group_by(voter_category, Q21) |>
  # Instead of counting, sum the weights
  summarize(weighted_count = sum(weight), .groups = "drop_last") |>
  # Calculate percentages based on weighted counts
  mutate(weighted_total = sum(weighted_count),
         weighted_percentage = weighted_count / weighted_total * 100) |>
  ungroup()

weighted_summary <- weighted_by_voter_category |>
  pivot_wider(
    id_cols = voter_category,
    names_from = Q21,
    values_from = weighted_percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

weighted_summary
```

**Explanation:** - Now we've calculated percentages by summing the
weights instead of simply counting people - Each respondent contributes
to the total based on their assigned weight value - This gives us
estimates that should better reflect the true population - Notice how
some percentages have changed from the unweighted analysis

## Part C: Comparing Differences

Let's create a comparison table to clearly see the differences between
weighted and unweighted results:

```{r category-comparison}
comparison <- unweighted_summary |>
  # Join the two tables to compare them
  inner_join(weighted_summary, by = "voter_category", suffix = c("_unweighted", "_weighted")) |>
  # Calculate the differences in percentage points
  mutate(
    yes_diff = `Yes (%)_weighted` - `Yes (%)_unweighted`,
    no_diff = `No (%)_weighted` - `No (%)_unweighted`,
    unsure_diff = `Unsure (%)_weighted` - `Unsure (%)_unweighted`
  ) |>
  # Select just the differences for our table
  select(voter_category, yes_diff, no_diff, unsure_diff) |>
  rename(
    "Yes (% point diff)" = yes_diff,
    "No (% point diff)" = no_diff,
    "Unsure (% point diff)" = unsure_diff
  )

comparison 
```

**Explanation:** -


This table shows how many percentage points the results changed after
applying weights - Positive numbers mean the weighted percentage is
higher; negative numbers mean it's lower - For example, after weighting,
the percentage of "rarely/never" voters saying "Yes" decreased by 3.6
percentage points - These differences might seem small, but in a close
election, even small shifts can be meaningful for reporting


**Answer These Questions:**

1.  Which voter category shows the biggest difference between weighted
    and unweighted results? **Write your answer here:** Voter category, rarely/never who voted yes to Q21 saw the biggest difference of 3.6 and 1.9 between the weighted and unweighted results. 

2.  How might these differences affect reporting on likely voter
    turnout? **Write your answer here:** the increase of 1.9 percentage points among those in the rarely/never voter category who said they would not vote, along with a 1.5-point rise among those unsure about voting in the 2020 election, suggests that people with a history of low participation are not going to turn out. As a result, reporting might say that habitual nonvoters are not expected to have a major impact on election results. 

3.  Why do you think these differences exist? What does this tell us
    about who might be over or under-represented in the sample? **Write
    your answer here:**  The differences tell us that the sample may overrepresent active voters because even the nonhabitual voters who said yes had a decrease and underrepresent those who are unlikely to vote as the unsure and no answers are the only ones with increases. These differences exist because active voters are more likely to respond to surveys and stay informed to say they'll vote while inactive voters are less likely to respond because they're less connected to what's happening. 
