---
title: "HW11: Education Level and Survey Weighting"
author: "Derek Willis/Rob Wells"
date: 10-28-2025
---

**"Princess Tatsi"**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# Introduction

In this assignment, you'll examine how survey weighting affects our understanding of voting intentions across different education levels, using the data we examined in class. You'll calculate both weighted and unweighted statistics, create visualizations, and reflect on the implications for reporting.

## The Data

The nonvoters dataset contains survey responses about voting intentions and behaviors from a national survey. The survey was conducted prior to an election and includes demographic information like education level.

```{r}
# Load the dataset
nonvoters_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/nonvoters_data.csv")

# Take a quick look at the data structure
glimpse(nonvoters_data)
```

### Key Variables

- `weight`: Survey weight assigned to each respondent
- `Q21`: Voting intention (1 = Yes, 2 = No, 3 = Unsure/Undecided)
- `educ`: Education level (College, Some college, High school or less)

## Task 1: Education Distribution

First, let's examine the distribution of education levels in our sample. **Replace "REPLACE_ME"** with the correct variable for education level.

```{r}

education_distribution <- nonvoters_data |>
  count(educ) |>
  mutate(percentage = n / sum(n) * 100)

education_distribution
```

## Task 2: Reflection Question

Why might education levels in survey samples often differ from the general population? What factors might cause certain education groups to be over or underrepresented?

**Write your ANSWER HERE** 

Survey samples don't always reflect the true population because some groups are less likely to take part due to less civic engagement or political stances This creates over- or under-representation in key characteristics like education. In this case, people with college degrees participated more in the survey, probably because they are more willing/able to respond than those without college education. Also because they might feel the most credibility to participate in the survey. 


## Task 3: Unweighted Analysis by Education

Now, let's calculate unweighted voting intentions by education level. This is what we would report if we didn't apply any weighting to our sample.

```{r}
# Calculate unweighted voting intentions by education
unweighted_by_education <- nonvoters_data |>
  # Filter out missing values
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  # Group by education and response
  group_by(educ, Q21) |>
  # Count responses
  summarize(count = n(), .groups = "drop_last") |>
  # Calculate percentages
  mutate(total = sum(count),
         percentage = count / total * 100) |>
  ungroup()

# Create a more readable format with voting intentions as columns
unweighted_educ_summary <- unweighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

unweighted_educ_summary
```

## Task 4: Reflection Question

Based just on this unweighted analysis, what headline might you write for a news story about education and voting intentions?

**Write your ANSWER HERE** Educational levels significantly impacts your likeliness to vote (now would need to test for significance, lol)

## Task 5: Weighted Analysis by Education

Next, let's apply survey weights to see how this changes our results. Instead of just counting responses, we'll sum the weights for each group. **Replace "REPLACE_ME"** with the appropriate weight variable

```{r weighted-by-education}

weighted_by_education <- nonvoters_data |>
  # Filter out missing values
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  # Group by education and response
  group_by(educ, Q21) |>
  # Sum the weights instead of counting
  summarize(weighted_count = sum(Q21), .groups = "drop_last") |>
  # Calculate weighted percentages
  mutate(weighted_total = sum(weighted_count),
         weighted_percentage = weighted_count / weighted_total * 100) |>
  ungroup()

# Create a more readable format
weighted_educ_summary <- weighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = weighted_percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

weighted_educ_summary
```

## Task 6: Reflection Questions

1. How did the percentages change after applying weights? Which education group showed the biggest changes?

**Write your ANSWER HERE** Groups that were underrepresented in the unweighted table saw an increase because we applied their weights. For example, voters whose education is high school or less and who said they don't plan to vote were underrepresented before at 14.6%, now their percentage is 21.1%. 

2. Why might the weighted results be considered more accurate than the unweighted results?

**Write your ANSWER HERE** The weighted results might be considered more accurate than the unweighted results because it accounts for gaps in population representation so the results have more validity. 

## Task 7: Comparison of Weighted vs. Unweighted Results

Let's create a direct comparison table to see the differences more clearly.

```{r}

comparison <- unweighted_educ_summary |>
  inner_join(weighted_educ_summary, by = "educ", suffix = c("_unweighted", "_weighted")) |>
  mutate(
    # Calculate the differences between weighted and unweighted percentages
    yes_diff = `Yes (%)_weighted` - `Yes (%)_unweighted`,
    no_diff = `No (%)_weighted` - `No (%)_unweighted`,
    unsure_diff = `Unsure (%)_weighted` - `Unsure (%)_unweighted`
  ) |>
  # Select just the columns we want to display
  select(educ, yes_diff, no_diff, unsure_diff) |>
  rename(
    "Education Level" = educ,
    "Yes (% point diff)" = yes_diff,
    "No (% point diff)" = no_diff,
    "Unsure (% point diff)" = unsure_diff
  )

comparison
```

## Task 8: Reflection Question

Which education group shows the largest differences between weighted and unweighted results?

**Write your ANSWER HERE** Voters with a high school or less education who answers yes to voting intentions saw the largest difference of -20.3% between weighted and unweighted results

## Task 9: Visualization

Visualizations can help us see the differences more clearly. Let's create a bar chart comparing weighted and unweighted "Yes" responses by education level. **Replace "REPLACE_ME"** with the correct variable name

```{r}
educ_viz_data <- bind_rows(
  # Unweighted data
  unweighted_by_education |> 
    filter(Q21 == 1) |>  # Only "Yes" responses (Q21=1)
    mutate(Type = "Unweighted") |>
    select(Type, educ, percentage),
  
  # Weighted data - 
  weighted_by_education |> 
    filter(Q21 == 1) |>  # Only "Yes" responses
    mutate(
      Type = "Weighted",
      percentage = weighted_percentage
    ) |>
    select(Type, educ, percentage)
)

# Create a grouped bar chart
ggplot(educ_viz_data, 
       aes(x = educ, y = percentage, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(
    title = "Weighted vs. Unweighted 'Yes' Responses by Education",
    subtitle = "Q21: Do you plan to vote in the November election?",
    y = "Percentage (%)",
    x = "Education Level"
  ) +
  scale_fill_manual(values = c("Unweighted" = "#619CFF", "Weighted" = "#F8766D")) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```


## Task 10: Final Reflection Questions

1. Why is it important for journalists to understand survey weighting when reporting on polls?

**Write your ANSWER HERE** It is important for journalists to understand survey weighting when reporting on polls because we never want to under represent or over represent groups of people as it will lead to inaccurate projections like in the 2020 election. 

It is important for journalists to understand survey weighting because it accounts for imbalances in who responds to a poll. Without weighting, some groups may be overrepresented while others are underrepresented. This can distort results, leading to misleading conclusions. Proper interpretation of weighted data helps to avoid overemphasing trends, mischaracterizing public opinion, or repeating errors seen in past elections such as the 2020 election where unweighted or poorly weighted samples contributed to inaccurate projections.

2. How might the differences between weighted and unweighted results affect how you would report on this data? 

**Write your ANSWER HERE** I would report on how college education impacts participation in politics. with a focus on the weight increase for voters who are high school graduates or completed some college, especially those who answered unsure (increase of 13.8 for high school graduates and 10.1 increase for some college) or no (increase of 6.496432 for high school graduates and increase of 4.816512 for some college ) to their voting intentions. I would also explain that  differences likely reflect underlying demographic participation patterns rather than genuine differences in political interest. To me, this signals a confounding variable affecting the participation of those who are not college graduates. 

3. What additional information would you want to know about how the weights were calculated before using this data in a news story?
**Write your ANSWER HERE** I would want to know the source of population data for which the survey was weighted on, the weighing methodlogy used. For example, if it was iterative proportional fitting aka raking, i would want to know what set of variables where used to adjust the weight for each case. 
