---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)

```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.
```{r}
library(dplyr)
```

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```

```{r}
install.packages("dplyr")
```

```{r}
library("tidyr")
```


### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

The data is divided into 14 categories of crime reports in PG County. Theft generally conflates the data, the first four categories, excluding accidents, are all theft related. The median would be necessary to get a real sense of the data, because one category has dominated the data set. From a reporting standpoint, theft-related incidents occur so frequently and can overshadow other categories in summaries. This makes the data harder to interpret, since the high volume of thefts inflates averages and masks patterns in less common but potentially more serious offenses.

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |>
  mutate(
    total_incidents = rowSums(across(Jan:Dec), na.rm = TRUE),
    average_incidents = rowMeans(across(Jan:Dec), na.rm = TRUE)
  )
```


### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

Accident, auto stolen, theft from auto, accident with impound, and theft are the five categories with the highest average of incidents per month. Robbery residential, homicide, B & E Other, sex offense, and assualt with weapon have the lowest average of incidents per month. It tells me that most crime reported and responded to is auto and theft related, with a healthy amount of stolen vechile's recovered. 


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime |>
  mutate(
    average_incidents = rowMeans(across(Jan:Dec), na.rm = TRUE),
    jun_change = (Jun - average_incidents) / average_incidents * 100,
    jul_change = (Jul - average_incidents) / average_incidents * 100,
    aug_change = (Aug - average_incidents) / average_incidents * 100
  )
  
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

the theory that crime is more likely to occur in the summer months is not strongly supported by my data because though some categories can rise slightly, the overall pattern does not show a systematic spike in summer crime. Homicide, though it one of the least occuring crimes but one of the most severe of offenses, sees the biggest spike in the summer months. Also rise in assualt shooting. 


### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

*Lede
As overall crime remains steady throughout the year in Prince George's Count, violent crimes such as Homicide and Assualt shootings see their sharpest rise during the summer months, according to data analysts. 

*Nutgraf
An analysis of monthly crime data shows that while most categories of crime in Prince George’s County do not display a consistent summer spike, serious offenses such as homicides and assault shootings do increase noticeably in the later summer months of July and August. Homicides, which are among the least common but most severe crimes, show the largest percentage increase during the summer. This can point to a seasonal vulnerability that affects community safety. These findings suggest that while crime overall may not be strongly tied to warmer weather, targeted prevention strategies during the summer could 
necessary to maintain public safety and well-being. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
