---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: <https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv>. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv") 
```

```{r}
glimpse(wh_salaries)
```


### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

Each row reflects the salary of a white house employee and detailee, which is a federal employee temporarily assigned to another agency for a specific project. There are five columns for name, status, salary, and title. From a reporting standpoint, some salaries amounting to zero stands out to me. This could cause issues with data analysis because it conflates the mean average. Glancing at the data, there are about 5 different salary amounts, and who gets what amount looks random and not tied to rank, maybe department. Also, every column is a character and R cannot find the mean/median of non-numeric values.  

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |>
  #mutate(salary = as.numeric(salary)) |>
  summarise(
    mean_salary = mean(salary),
    median_salary = median(salary)
  )
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |>
  summarise(
    mean_salary = mean(salary),
    median_salary = median(salary),
    max_salary = max(salary),
    min_salary = min(salary)
  )
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The max salary was 251,258 but my code gave me 99,315. Even when i order the data set 99,315 is the top number, I had to scroll to see the true maximum salary. The minimum salary is correct, some people were paid zero dollars, according to the data. I can't determine if they're outliers because I do not have a median and mean to compare. The problem I was trying to figure out, and what I need help with, is converting a character data type to a numeric data type. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?
Out of the 11 people paid zero dollars, 7 were detailee and 4 employees of the white house. The highest paid person also a detailee. My best story idea would investigate why this title has inconsistent compensation. Are the zero-dollar detailees volunteers from other agencies? Or personnel whose salaries are covered elsewhere? Or unpaid White House interns holding the "detailee" title? What specific high-level role warrants the top salary for a detailee, and how does that role contrast with the unpaid positions? So far, I can see that BARRANS, MICHELLE M., paid 251,258, served as Associate Counsel. The other detailees paid zero dollars also served advisory roles. So the next steps would be talking to people familiar with how the White house staffing works to get answers to these questions. 
